The Role of Ice Compositions and Morphology for Snowlines and the C/N/O Ratios in Active Disks

### 2016-03-10 tex/draft2/ms_edited.tex
update

\begin{abstract}
The elemental compositions of planets define their chemistry, and could potentially be used as beacons for their formation location if the elemental gas and grain ratios of planet birth environments, i.e. protoplanetary disks, are well understood. In disks, the ratios of volatile elements, such as C/O and N/O, are regulated by the abundance of the main C, N, O carriers, their ice binding environment, and the presence of snowlines of major volatiles at different distances from the central star. We explore the effects of dynamical processes, molecular compositions and abundances, and the ice morphology of dust grains in disks on the snowline locations of the main C, O and N carriers, and their consequences for the C/N/O ratios in gas and dust throughout the disk. 
%We find that radial drift and accretion alone can reduce the snowline radii of the main C, O and N carriers, i.e. H$_2$O, CO$_2$, CO and N$_2$, by 40-60\% compared to static disks. 
The gas-phase N/O ratio enhancement in the outer disk (exterior to the H$_2$O snowline) exceeds the C/O ratio enhancement for all reasonable volatile compositions. Ice morphology and disk dynamics individually change the snowline location of N$_2$, the main nitrogen carrier, by a factor of 2-3, and when considered together the range of possible N$_2$ snowline locations is $\sim$11-$\sim$79 AU in a standard disk model. The CO snowline location is similarly uncertain. This snowline location uncertainty limits the utility of C/N/O compositional ratios in planets to trace their formation zone. Observations that anchor snowline locations at different stages of planet formation are therefore key.

@@

Due to disk dynamics and ice morphology, the disk regions with highly elevated gas phase N/O and C/O ratios are uncertain and may span tens of AU. Both ice morphologies discussed in this study, pure and water dominated ices, are plausible in protoplanetary disks and depend on whether H$_2$O and CO ices formed on similar timescales or successively (e.g., \citealt{garrod11}). Observations of protostellar cores show that a large fraction of CO is bound in a pure ice multilayer \citep{pontoppidan03}, but theoretical models also suggest an icy mantle structure where CO resides on a H$_2$O ice layer (e.g., \citealt{collings03}). One can also imagine a scenario where CO is in a water binding environment and N$_2$ is not. This could be attributed to the fact that H$_2$O may bind preferentially to CO than N$_2$, since both H$_2$O and CO are polar molecules while N$_2$ is not. It is also possible for N$_2$ ices to form later than CO (e.g., \citealt{pagani12}), and thus be deposited on the outer layers of the icy mantles which are typically water poor (e.g., \citealt{garrod11}). Detections of CO and N$_2$ snowlines in the same disk could potentially verify this scenario, since CO and N$_2$ desorb at substantially different temperatures, and therefore locations, if the former is water dominated and the latter is in a pure ice environment. The impact of the ice environment on the snowline location is much smaller in the case of CO$_2$ and NH$_3$, as their binding energies and behavior are closer to that of H$_2$O. No detailed measurements for the CH$_4$ binding energy in a water environment exist so far, but due to its low desorption temperature a similar behavior to that of CO and N$_2$ would be expected. While the presence of some carbon in the form of CH$_4$ only modestly affects our results, CH$_4$ may become important in disks where a large fraction of the CO abundance has been converted into hydrocarbons (e.g., \citealt{du15}).

Changes in stellar luminosity (e.g., \citealt{kennedy06}) and gas mass accretion rate (e.g., \citealt{chambers09}), as well as the evolution of icy dust particles due to grain growth and fragmentation (e.g., \citealt{birnstiel12}), may introduce additional uncertainties in the snowline locations, and thus the C/N/O ratios. Given the number of uncertainties in snowline locations, detections of snowlines in a sample of disks at different evolutionary stages are needed to provide observational constraints. Current detections include the H$_2$O snowline in TW Hya \citep{zhang13} and the CO snowline in TW Hya \citep{qi13} and HD 163296 \citep{qi15}. However, more snowline detections in disks of different ages are expected in future ALMA cycles.

>>>>

\begin{abstract}
The elemental compositions of planets define their chemistry, and could potentially be used as beacons for their formation location if the elemental gas and grain ratios of planet birth environments, i.e. protoplanetary disks, are well understood. In disks, the ratios of volatile elements, such as C/O and N/O, are regulated by the abundance of the main C, N, O carriers, their ice binding environment, and the presence of snowlines of major volatiles at different distances from the central star. We explore the effects of dynamical processes, molecular compositions and abundances, and the ice morphology of dust grains in disks on the snowline locations of the main C, O and N carriers, and their consequences for the C/N/O ratios in gas and dust throughout the disk. The gas-phase N/O ratio enhancement in the outer disk (exterior to the H$_2$O snowline) exceeds the C/O ratio enhancement for all reasonable volatile compositions. Ice morphology and disk dynamics individually change the snowline location of N$_2$, the main nitrogen carrier, by a factor of 2-3, and when considered together the range of possible N$_2$ snowline locations is $\sim$11-$\sim$79 AU in a standard disk model. The CO snowline location is similarly uncertain. This snowline location uncertainty limits the utility of C/N/O compositional ratios in planets to trace their formation zone. Observations that anchor snowline locations at different stages of planet formation are therefore key.

@@

Due to disk dynamics and ice morphology, the locations of the CO and N$_2$ snowlines, and thus the disk regions with highly elevated gas phase N/O and C/O ratios, are uncertain and may span tens of AU. Both ice morphologies discussed in this study, pure and water dominated ices, are plausible in protoplanetary disks and depend on whether H$_2$O and CO ices formed on similar timescales or successively (e.g., \citealt{garrod11}). Observations of protostellar cores show that a large fraction of CO is bound in a pure ice multilayer \citep{pontoppidan03}, but theoretical models also suggest an icy mantle structure where CO resides on a H$_2$O ice layer (e.g., \citealt{collings03}). One can also imagine a scenario where CO is in a water binding environment and N$_2$ is not. This could be attributed to the fact that H$_2$O may bind preferentially to CO than N$_2$, since both H$_2$O and CO are polar molecules while N$_2$ is not. It is also possible for N$_2$ ices to form later than CO (e.g., \citealt{pagani12}), and thus be deposited on the outer layers of the icy mantles which are typically water poor (e.g., \citealt{garrod11}). The impact of the ice environment on the snowline location is much smaller in the case of CO$_2$ and NH$_3$, as their binding energies and behavior are closer to that of H$_2$O. No detailed measurements for the CH$_4$ binding energy in a water environment exist so far, but due to its low desorption temperature a similar behavior to that of CO and N$_2$ would be expected. While the presence of some carbon in the form of CH$_4$ only modestly affects our results, CH$_4$ may become important in disks where a large fraction of the CO abundance has been converted into hydrocarbons (e.g., \citealt{du15}).

Changes in stellar luminosity (e.g., \citealt{kennedy06}) and gas mass accretion rate (e.g., \citealt{chambers09}), as well as the evolution of icy dust particles due to grain growth and fragmentation (e.g., \citealt{birnstiel12}), may introduce additional uncertainties in the snowline locations, and thus the C/N/O ratios. Given the number of uncertainties in snowline locations, detections of snowlines in a sample of disks at different evolutionary stages are needed to provide observational constraints. The uncertainties in snowline locations caused by disk dynamics, ice morphology, and other effects outlined above can be resolved in extreme cases, such as a detection of a CO snowline at a temperature corresponding to pure CO ice desorption in a static disk (e.g., \citealt{qi13} at $\sim$ 17 K). In other cases, however, the degeneracies in snowline locations cannot be broken. For example, the CO snowline in HD 163296 is at a higher temperature of $\sim$ 25 K \citep{qi15}, which could be caused either by CO being in a water dominated environment or by dynamical effects that push the CO snowline inward. Detections of both the CO and N$_2$ snowlines in the same disk could potentially break this degeneracy, since CO and N$_2$ desorb at substantially different temperatures, and therefore locations, if the former is water dominated and the latter is in a pure ice environment.

%Current detections include the H$_2$O snowline in TW Hya \citep{zhang13} and the CO snowline in TW Hya \citep{qi13} and HD 163296 \citep{qi15}. More snowline detections in disks of different ages are expected in future ALMA cycles.  Snowline observations may also be used to constrain the degree of turbulence in protoplanetary disks \citep{owen14}. We note, however, that observations of snowlines may not be enough to break the degeneracies in snowline locations produced by disk dynamics, ice morphology, and other effects outlined above. For example, a detection of the CO snowline at a temperature higher than the assumed desorption temperature in a static disk with CO as pure ice ($\sim$17 K, e.g. \citealt{qi13}) could be caused either by CO being in a water dominated environment or by dynamical effects that push the CO snowline inward. This degeneracy can be broken in extreme cases --- for our fiducial disk model, this implies a detection of a CO snowline close to $\sim$9 AU or $\sim$61 AU (see Section \ref{sec:dynamic}).
